A Large-scale Study of Polymorphism in R

### 2019-03-17 paper/main.tex

\title{A large-scale study of polymorphism in R}

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in over X millions
of lines of R code spread among some 7500 packages, written over a period of over 20 years by over 10000
programmers.  We perform a dynamic analysis, leveraging tests and use-cases,
to determine the level of polymorphism that is present in the code. We do
this for several potential notions of types. Our result suggest that
polymorphism is important in some key parts of the system but that
relatively simple type annotations could be used to capture most of the
interesting cases.
\end{abstract}

@@

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit most users of a language, we need to
understand the language we are working with as well as the real-world needs
it answers. Often, we, as researchers, can appeal to our intuition as many
languages are intended to be general purpose and appeal to users with some 
training. Unfortunately, these intuitions don't always apply to domain-specific
languages.

A number of programming languages have gained popularity for data science
tasks, they include MATLAB, Python, Julia and R.

These languages are all dynamically typed.

This is because they emphasize ease of programming, interactive data
exploration, and light-weight syntax.

To facilitate programming these language offer all kinds of short cuts and
facilities that increase the likelihood that code will ``just work''.

For example abbreviation of name arguments and automatic conversions.

But invariably some applications written in those languages survive past the
prototype stage.

The results obtained are published and even sometimes drive policy decisions.
\AT{There's a cool stat in the blog I linked which says that a lot of the Stack
Overflow traffic R gets is from government officials.}

Given that code almost never stops, and that any computation is likely to
deliver a bunch of numbers, how can we trust those numbers?

Years of work in programming languages suggest that one cost efficient way
to catch some errors is to add types to a language.

In the last decade much attention has been given to the incremental addition
of types. Successes like Hack, TypeScript and Racket suggest that this may
be feasible for wide spectrum languages.

R is a languages written by and for \textit{statisticians}.  It traces its
roots to the S language~\cite{becker1988new}, developed by Richard Becker, 
John Chambers, and Allan Wilks.  R programmers don't write big
servers or software systems; instead, they write and modify small data
processing pipelines as a means to facilitate interaction with digital data.
This leads to ``front-line'' R code (i.e., code written for users, not
programmers) not being made available on typical platforms like Github.
\AT{Maybe front-line code isn't a good word, is there a good way to describe
  non-library code?}  Put differently, R is essentially a really big
domain-specific language.

The design of R is also rife with questionable decisions.  For example, the
language is lazy \textit{and} side-effecting, which one could describe as an
antagonistic pair of features.  In addition, data structures must be
entirely copied when being modified, which appears to be far from ideal in a
language meant to process (thus, modify) big data.

In spite of these flaws, the language is incredibly popular.
\AT{We'll want to cite something here, maybe the blog I linked?}

>>>>

\title{A Large-scale Study of Polymorphism in R}

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in over X millions
of lines of R code spread among some 7,500 packages, written over a period of
over 20 years by over 10,000 programmers.  We perform a dynamic analysis,
leveraging tests and use-cases, to determine the level of polymorphism that
is present in the code. We do this for several potential notions of
types. Our result suggest that polymorphism is important in some key parts
of the system but that relatively simple type annotations could be used to
capture most of the interesting cases.
\end{abstract}

@@

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit most users, we need to understand the
language we are working with as well as the real-world needs it
answers. Often, we, as researchers, can appeal to our intuition as many
languages are intended to be general purpose and appeal to users with some
computer science training. Unfortunately, these intuitions don't always
apply to domain-specific languages, languages designed for and by a specific
group of users to solve very specific needs. This is the case of the data
science language R.

R and it's ancestor S are languages designed, implemented and maintained by
statisticians. Originally they were designed as glue languages, languages
that would allow to read data into vectors and call statistical routines
written in Fortran. Over three decades, the languages became widely used
across many fields of science and in industry for data analysis and data
visualization; with time additional features were added.  Modern R, as a
linguistic object of study, is fascinating. It is a vectorized, dynamically
typed, lazy functional language with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.

Many of the design decisions that gave us R were intended to foster an
interactive, exploratory, programming style. This includes, to name a few,
the lack of type annotations on variables and functions, the ability to use
syntactic shortcut, and the automatic conversion between data types.  While
these choices have led to a language that is surprisingly easy to use by
beginners --many data science programs do not teach the language itself but
simply introduce some of its key libraries-- they have also created a
language where almost all computations yield a numeric result and where
errors can go undetected.

One way to increase assurance in the results obtained when using R would be
to add type annotations to functions and variable declarations. These
annotations could then be used, either statically or (more likely)
dynamically, to catch mismatches between expected and provided data values.
The nature of R is such that it is unlikely to be ever fully statically
checked, furthermore end users may not be willing to write types when
carrying out exploratory programming tasks. So, we are looking for an
optional type system that would allow us to capture as much of behavior of
library functions as possible while remaining easy to understand for
end-users and library developers alike.
